Solving the Shallow Water Equations\\ using Finite Volumes and Lax-Friedrichs

### 2017-12-21 paper/paper.tex
Conclusion and bibliography

\input{abstract}

\begin{keywords}
	Shallow water equations, Finite volumes, Lax-Friedrichs method, Rusanov, artificial viscosity, Julia
\end{keywords}

@@

The shallow water equations are a system of hyperbolic partial differential equations which are effective at describing the behavior of tsunamis and overland flooding. They are derived from the general Navier-Stokes equations by making the following assumptions:

@@

This latter problem may be resolved quickly, thanks to Julia's type system.

@@

Finite volumes are a family of numerical methods which are well suited for computing an approximate solution to hyperbolic systems such as the shallow water equations. They have two main advantages over other methods, such as finite differences and finite elements. Firstly, they automatically conserve physical quantities $q$. Secondly, they can capture discontinuities, such as shocks, which is considerably simpler than explicitly tracking them. The main downside is that the profile of the shock will be smeared by the spatial discretization.

@@

Ideally, the flux function can be calculated using only the state in the adjacent cells, which implies a fully discrete method:

@@

This yields a forward-time, centered-space finite difference scheme. Unfortunately a von Neumann analysis shows that it is always unstable. The Lax-Friedrichs method simply modifies this scheme by changing the time difference to use a value of $Q_i^n$ averaged from the neighboring cells:

@@

\section{Comparison of results}
\begin{figure}[h]%
 	\begin{center}%
 		\includegraphics[scale=0.6]{../plots/lxf_llxf_breakingdam.png}%
 		\caption{Baum}\label{fig:baum}%
 	\end{center}%
\end{figure}
\begin{figure}[h]%
 	\begin{center}%
 		\includegraphics[scale=0.6]{../plots/lxf_llxf_gaussian.png}%
 		\caption{Baum}\label{fig:baum}%
 	\end{center}%
\end{figure}

\begin{itemize}
\item For Lax-Friedrich, $\mu$ affects dissipation
\item For Local-Lax-Friedrich, $\mu$ affects dissipation much less
\item Lax-Friedrich approaches Local-Lax-Friedrich as $\mu \to 1.0$
\item Both methods are unstable for $\mu > 1.0$ due to CFL violation
\end{itemize}

@@

@@

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,references}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{ref:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}

>>>>

%\input{abstract}

%\begin{keywords}
%	Shallow water equations, Finite volumes, Lax-Friedrichs method, Rusanov, artificial viscosity, Julia
%\end{keywords}

@@

The shallow water equations are a system of hyperbolic partial differential equations which are effective at describing the behavior of tsunamis and overland flows.\cite{delestre:hal-00932234} They are derived from the general Navier-Stokes equations by making the following assumptions:

@@

This latter problem may be resolved quickly, thanks to Julia's type system.\cite{doi:10.1137/141000671}

@@

Finite volumes are a family of numerical methods which are well suited for computing an approximate solution to hyperbolic systems such as the shallow water equations.\cite{leveque2002finite} They have two main advantages over other methods, such as finite differences and finite elements. Firstly, they automatically conserve physical quantities $q$. Secondly, they can capture discontinuities, such as shocks, which is considerably simpler than explicitly tracking them. The main downside is that the profile of the shock will be smeared by the spatial discretization.

@@

Ideally, the flux function can be calculated using only the state in the adjacent cells, which yields a fully discrete method:

@@

This yields a forward-time, centered-space finite difference scheme. Unfortunately a von Neumann analysis shows that it is always unstable.\cite{strang2007computational} The Lax-Friedrichs method simply modifies this scheme by changing the time difference to use a value of $Q_i^n$ averaged from the neighboring cells:

@@

The Local Lax-Friedrichs method does not entirely break the coupling between dissipation and $\mu$, but it strongly reduces it, as depicted in Figure~\ref{fig:llxf_breakingdam_mu}. This experiment used the exact same parameters as Figure~\ref{fig:lxf_breakingdam_mu}, only for LLxF. 
\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/llxf_breakingdam_mu.png}%
        \caption{Baum}\label{fig:llxf_breakingdam_mu}%
    \end{center}%
\end{figure}

\section{LxF vs LLxF Comparison}

\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_breakingdam.png}%
        \caption{Baum}\label{fig:cmp_breakingdam}%
    \end{center}%
\end{figure}
\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_gaussian.png}%
        \caption{Baum}\label{fig:cmp_gaussian}%
    \end{center}%
\end{figure}

A final round of experiments were conducted to compare Lax-Friedrichs to Local Lax Friedrichs. These indicate that Local Lax-Friedrichs is decidedly less dissipative and more accurate. Figure~\ref{fig:cmp_breakingdam} shows the results of running LLxF with a middling $\mu=0.4$ against LxF with different $\mu$ for the breaking dam scenario.

At $\mu=0.4$, the LLxF method performs only slightly worse than LxF at $\mu=1.0$. For any given choice of $\mu$, the LLxF reaches a better solution in the same number of timesteps. Unexpectedly, the quality of the solutions converge as $\mu \to 1.0$. On the other hand, LLxF requires additional work to calculate $a(x)$ for each cell, though the amount is small and linear in the number of cells.

The stairstep oscillations observed in the breaking dam case for LxF are not present for LLxf. This is presumably because the value of $Q_i^n$ enters into the calculation of $Q_i^{n+1}$, albeit rather indirectly, via the max wavespeed $\lambda_{max}$ and then the cell-wise flux $F_{i+1/2}^n$.

@@

The Lax-Friedrichs method is a simple finite volume technique for solving the shallow water equations. Starting from the fundamentals, this paper built up a reference implementation which leverages features of Julia to achieve terseness and clarity not available in languages such as Matlab or C++. An abstract finite volume skeleton allows different numerical methods to be swapped in. Two variants were implemented, Lax-Friedrichs and Local Lax-Friedrichs. The former suffers from excessive numerical viscosity, which the latter largely remedies. The causes and consequences of this viscosity were discussed with respect to two model problems, a Gaussian perturbation and a breaking dam. It is the author's hope that this document may serve as a concise guide to students exploring this material in the future.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,../references}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{ref:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}
